On the Minimum Core Mass for Giant Planet Formation

### 2013-06-14 tex/draft/draft2.tex
current paper versions, plus figures for paper II

%\newcommand{\delad}{\nabla_{\rm ad}}
%\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\co}{_{\rm c}}
\newcommand{\di}{_{\rm o}}
\newcommand{\cb}{_{\rm RCB}}
\newcommand{\surf}{_M}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1\oplus} }

@@

\title{On the Minimum Core Mass for Giant Planet Formation through Kelvin-Helmholtz Contraction}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}

\begin{abstract}
The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  ``critical core mass''.  We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars. Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas. Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.

@@

Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc. \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.

Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected. These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.

Forming giant planets at wide separations in the disk poses theoretical challenges. On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc. \citealt{rafikov05}). On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs). I would therefore be easier if giant planets could form from smaller cores which would need less time to grow.

In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs. In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere. This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc. In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope. The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.

In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible. As a result, the protoplanetary core has a fixed mass. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences. The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''.

We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time. We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.

This paper is organized as follows. In section \ref{sec2} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere. In section \ref{analytic}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model. We describe our results in section \ref{KH}, and determined the critical core mass for planet formation during the life time of the protoplanetary disk in section \ref{critical}.  We discuss our results in section \ref{discussion} and summarize our findings in section \ref{conclusions}.

@@

In this section we derive the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk in the absence of planetesimal accretion. In section \ref{model} we describe the assumptions of our model. Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere.  In section \ref{BCs} we describe discuss our boundary conditions and the properties of our assumed protoplanetary disk, and review relevant length scales.  and the boundary conditions that we use. In section \ref{cooling} we develop a model to estimate the rate at which the atmosphere cools. Finally, we combine these calculations into a quasi-static model for atmospheric evolution in section \ref{twolayer}.

@@

\begin{enumerate}
\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk, as mentioned in section. The two regions are separated by the Schwarzschild criterion for convective instability (see section \ref{struct}). We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R_{RCB}$.
\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion. 
\item The luminosity is assumed to be constant throughout the outer radiative region. Since the structure of the convective zone is independent of its luminosity, we do not need to make additional assumptions about the luminosity in this region.
\item The envelope evolves through stages of quasi-static equilibrium.
\end{enumerate}

@@

In order for equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$. In our study, we assume an ideal gas polytropic EOS

@@

\noindent with $K$ the adiabatic constant. The adiabatic gradient is $\delad=2/5$ for a monatomic gas and $\delad=2/7$ for a diatomic gas. Additionally, the ideal gas law relates the pressure, temperature and density through

@@

\noindent with $\mathcal{R}$ defined in section \ref{model}. The specific gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$. $\mu$ is determined by the relative abundance of hydrogen and helium in the atmosphere. In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index. We show in section \ref{critcore} that this assumption does not affect the qualitative behavior of our solutions.

@@

\begin{equation}
\label{eq:opacitylaw}
\kappa=\kappa_0 \Big(\frac{P}{P_d}\Big)^{\alpha} \Big(\frac{T}{T_d}\Big)^{\beta},
\end{equation}

\noindent with $\alpha$ and $\beta$ constants, and $\kappa_0=2 F_{\kappa}$. To estimate $\alpha$ and $\beta$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $F_{\kappa}=1$ (i.e., $\kappa_0=2$). We note that these values are valid only for low disk temperatures: $T_d< \sim 100 K$. As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 10$ AU).

@@

We first define several characteristic length scales that are important for this problem. \\

@@

\begin{equation}
\label{eq:rc}
R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
\end{equation}

\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\

@@

\begin{equation}
\label{eq:RB}
R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
\end{equation}

\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. \\

\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by

\begin{equation}
\label{eq:RHill}
R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
\end{equation}

\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\

@@

\begin{equation}
\label{eq:H}
H \equiv \frac{c_s}{\Omega},
\end{equation}

\noindent where $\Omega$ is the Keplerian angular velocity. The disk scale height is a measure of the thickness of the protoplanetary disk. Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as

\begin{equation}
\label{eq:H2}
H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
\end{equation}

For small planet masses, $R_B<R_H<H$. As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$. Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.

We assume a solid core of fixed mass $M_c$ with a radius given by equation (\ref{eq:rc}). We choose $\rho_c=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}). Furthermore, we assume the atmosphere extends out to the boundary of the Hill sphere, defined by equation (\ref{eq:RHill}). If the Bondi radius (defined in equation (\ref{eq:RB})) is within the Hill radius, perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the core. If, on the other hand, $R_B>R_H$, then the Hill radius is the more relevant truncation radius since material cannot be gravitationally bound to the protoplanet outside the Hill sphere. As such, it is always safe to choose the Hill radius as the outer boundary.

As stated in section \ref{model}, the atmosphere matches smoothly onto the disk at the outer boundary. The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_H)=T_d$ and $P(R_H)=P_d$. As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma_d&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T_d &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
\end{subeqnarray}

\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN). In this study, we assume $F_{\Sigma}=F_T=1$. The resulting mid-plane pressure is given by

\begin{equation}
\label{eq:Pd}
P_d=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyne cm}^{-2}
\end{equation}

\noindent for a molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We take $m_*=1$.

@@

We first describe the general ways in which planetary Klevin-Helmholtz contraction differs from the stellar case.

An isolated sphere, such as a star, satisfies a simple global energy equation:

@@

A protoplanetary atmosphere that is not isolated, but is embedded in a gas disk, satisfies a more complex cooling equation:

\begin{equation}
\label{eq:coolingglobal}
L=L_c+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}
\end{equation}

@@

\begin{equation}
\label{eq:Eg}
E_G=-\int_{M_c}^M \frac{G m}{r} dm,
\end{equation}

@@

\begin{equation}
\label{eq:U}
U=\int_{M_c}^M u dm,
\end{equation}

\noindent where $u$ is the internal energy per mass. We now explain the additional terms in the cooling equation (\ref{eq:coolingglobal}). $L_c$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay. $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a surface mass element.

The global cooling equation can be derived starting from the local cooling balance. Integrating the local energy equation and (\ref{eq:structd}) from the core to the surface and using equation (\ref{eq:structb}) yields

@@

We consider a two-layer atmosphere structure model as presented in section \ref{model}. We approximate the outer radiative zone of the atmosphere as nearly isothermal. Moreover, we do not take into account the self-gravity of the atmosphere in either region. In the next subsections we derive the structure and cooling behavior of such an atmosphere, deferring some of the calculations to \App{sec:analytic}.

@@

We now consider the structure of a low mass (non self-gravitating) isothermal atmosphere.  We assume the atmosphere matches onto a constant background density, $\rho\di$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$.  The resulting density profile is
\begin{equation} \label{eq:rhoiso}
\rho = \rho_{\rm d} \exp \left({R_B \over r} - {1 \over n_B} \right) \approx   \rho_{\rm{d}} \exp \left(R_B \over r  \right),
\end{equation} 
where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.  However the choice of boundary condition does have an order unity effect on the density near the Bondi radius. In our study, $n_{\rm{fit}}=R_{\rm B}/R_{\rm H}$, when the Bondi radius is larger than the Hill radius, and $n_{\rm fit}=1$ when the opposite is true.

The mass of the atmosphere is typically determined by integrating the density profile from the core to the Bondi radius.  Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.  In this case
\begin{equation} \label{eq:MatmISO}
M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
\end{equation} 
This result is the leading order term in a series expansion.  Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.

@@

%\subsection{Temperature Contrast at Convective Boundary}
We now show that the temperature contrast between the radiative-convective boundary $T_{\rm RCB}$ and the ambient disk $T_{\rm d}$, is modest.  From equation (\ref{eq:delrad}), we express the radiative lapse rate
\begin{equation}\label{eq:delradan}
\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P_{\rm d})^{1+\alpha} \over (T/T_{\rm d})^{4-\beta}}
\end{equation}

\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}), and $\nabla \di$ is the radiative temperature gradient at the disk:

\begin{equation}
\label{eq:delo}
\nabla \di \equiv \frac{3 \kappa_0 P_{\rm d}}{64 \pi G M \sigma T_{\rm d}^4} L.
\end{equation}

>>>>

%\newcommand{\delad}{\nabla_{\rm ad}}
%\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\co}{_{\rm c}}
\newcommand{\di}{_{\rm d}}
\newcommand{\cb}{_{\rm RCB}}
\newcommand{\surf}{_M}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1\oplus} }

@@

\title{On the Minimum Core Mass for Giant Planet Formation}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}

\begin{abstract}
The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. This core mass is typically quoted to be around $10 M_{\oplus}$. In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  \textit{critical core mass}.  We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars. Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas. Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. %We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.

@@

Current theories of giant planet formation postulate that these planets form either through core accretion (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), where solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.

Standard core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. These studies consider that the planet atmosphere is in a steady state at all times, where all the incoming energy from planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.  Standard calculations find the critical core mass to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).

Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges. On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (e.g., \citealt{rafikov05}). On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the gas in the disk (e.g., \citealt{rafikov06}). It would therefore be easier if giant planets could form from smaller cores which need less time to grow.

In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs. In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere. This reduction can arise, for example, due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards. In this situation, the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion. The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.

In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible. As a result, the protoplanetary core has a fixed mass. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences. The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the \textit{critical core mass}.

We build a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time. We aim to find the critical core mass for a giant planet to form before disk dissipation.

This paper is organized as follows. In section \ref{sec2} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere. In section \ref{analytic}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model. We describe our numerical results in section \ref{KH}, and determine the critical core mass for planet formation during the life time of the protoplanetary disk in section \ref{critical}.  We discuss some of the approximations of our model in section \ref{neglected} and summarize our findings in section \ref{conclusions}.

@@

In this section we derive the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk in the absence of planetesimal accretion. In section \ref{model} we describe the assumptions of our model. Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere.  In section \ref{BCs} we discuss our boundary conditions and the properties of our assumed protoplanetary disk, and review relevant length scales.  In section \ref{cooling} we develop a model to estimate the rate at which the atmosphere cools. %Finally, we combine these calculations into a quasi-static model for atmospheric evolution in section \ref{twolayer}.

@@

\begin{enumerate}
\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative region that matches smoothly on to the protoplanetary disk. The two regions are separated by the Schwarzschild criterion for convective instability (see section \ref{struct}). We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R\cb$.
\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction, while planetesimal accretion is negligible. As such, the core no longer grows and has a fixed mass.
\item The luminosity is assumed to be constant throughout the outer radiative region. Since the structure of the convective zone is independent of its luminosity, we do not need to make additional assumptions about the luminosity in this region.
\item The envelope evolves through stages of quasi-static equilibrium.
\end{enumerate}

@@

In order for the equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$. In our study, we assume an ideal gas polytropic EOS

@@

\noindent with $K$ the adiabatic constant, determined by $\delad$. The adiabatic gradient is $\delad=2/5$ for a monatomic gas and $\delad=2/7$ for a diatomic gas. Additionally, the ideal gas law relates the pressure, temperature and density through

@@

\noindent where  $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. The gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$, which is determined by the relative abundance of hydrogen and helium in the atmosphere. In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index. We show in section \ref{critcore} that this assumption does not affect the qualitative behavior of our solutions.

@@

\begin{equation}
\label{eq:opacitylaw}
\kappa=\kappa_0 \Big(\frac{P}{P_{\rm{ref}}}\Big)^{\alpha} \Big(\frac{T}{T_{\rm{ref}}}\Big)^{\beta},
\end{equation}

\noindent with $\alpha$ and $\beta$ constants, $\kappa_0=2 F_{\kappa}$, and $T_{\rm{ref}}$ and $P_{\rm{ref}}$ reference temperature and pressure, respectively. To estimate $\alpha$ and $\beta$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$, $T_{\rm{ref}}=100$ K, and $F_{\kappa}=1$ (i.e., $\kappa_0=2$). We note that these values are valid only for low disk temperatures: $T_d \lesssim 100 K$. As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 5$ AU).

@@

In what follows we first define several characteristic length scales that are important for this problem. \\

@@

\begin{equation}
\label{eq:rc}
R\co \equiv \Big(\frac{3 M\co}{4 \pi \rho\co}\Big)^{1/3},
\end{equation}

\noindent with $M\co$ and $\rho\co$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\

@@

\begin{equation}
\label{eq:RB}
R_{\rm B} \equiv \frac{G M_{\rm p}}{c_s^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di},
\end{equation}

\noindent where $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_{\rm s}$ is the isothermal sound speed, defined as $c_{\rm s}=\Big(\frac{k_b T\di}{\mu m_p}\Big)^{1/2}$, and $\mathcal{R}$ is the reduced gas constant defined in section \ref{struct}.\\

\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and the tidal acceleration from the host star are the same. It is given by

\begin{equation}
\label{eq:RHill}
R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}
\end{equation}

\noindent Outside the Hill radius, the gravity of the planet is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\

@@

\begin{equation}
\label{eq:H}
H \equiv \frac{c_{\rm s}}{\Omega},
\end{equation}

\noindent where $\Omega$ is the Keplerian angular velocity. The disk scale height is a measure of the thickness of the protoplanetary disk. %Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as

%\begin{equation}
%\label{eq:H2}
%H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
%\end{equation}

For small planet masses, $R_{\rm B}<R_{\rm H}<H$. As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_{\rm H}<R_{\rm B}<H$. Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk. In our study, we only accept atmosphere solutions that are spherically symmetric.

We assume a solid core of fixed mass $M\co$ with a radius given by equation (\ref{eq:rc}). We choose $\rho\co=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}). Furthermore, we assume the atmosphere extends out to the boundary of the Hill sphere, defined by equation (\ref{eq:RHill}). If the Bondi radius (defined in equation \ref{eq:RB}) is within the Hill radius, perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the planet. If, on the other hand, $R_{\rm B}>R_{\rm H}$, then the Hill radius is the more relevant truncation radius since material cannot be gravitationally bound to the protoplanet outside the Hill sphere. As such, it is always safe to choose the Hill radius as the outer boundary. We note, however, that we define the planet mass as only the mass enclosed within the Bondi radius if $R_{\rm B}<R_{\rm H}$.%, i.e. $M_{\rm{tot}}=M\co+M_{\rm{atm}}=M(<\text{min}(R_{\rm B}, R_{\rm H}))$.

The atmosphere matches smoothly onto the disk at the outer boundary. The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_{\rm H})=T\di$ and $P(R_{\rm H})=P\di$. As a fiducial disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
\end{subeqnarray}

\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN). In this study, we assume $F_{\Sigma}=F_T=1$. The resulting mid-plane pressure and disk scale height are given by

\begin{equation}
\label{eq:Pd}
P\di=110  F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyn cm}^{-2}
\end{equation}
and
\begin{equation}
H=0.022 \sqrt{F_T} a^{9/7}\,\, \text{AU},
\end{equation}

\noindent respectively, for a molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We take $m_*=1$.

@@

We first describe the general ways in which planetary Klevin-Helmholtz contraction differs from the stellar case. An isolated sphere, such as a star, satisfies a simple global energy equation:

@@

A protoplanetary atmosphere that is not isolated, but rather embedded in a gas disk, satisfies a more complex cooling equation:

\begin{equation}
\label{eq:coolingglobal}
L=L\co+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}
\end{equation}

@@

\begin{equation}
\label{eq:Eg}
E_G=-\int_{M\co}^M \frac{G m}{r} dm,
\end{equation}

@@

\begin{equation}
\label{eq:U}
U=\int_{M\co}^M u dm,
\end{equation}

\noindent where $u$ is the internal energy per mass. We now explain the additional terms in the cooling equation (\ref{eq:coolingglobal}). $L\co$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay. $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a mass element as the atmosphere contracts.

The global cooling equation (\ref{eq:coolingglobal}) can be derived starting from the local cooling balance. Integrating the local energy equation and (\ref{eq:structd}) from the core to the surface and using equation (\ref{eq:structb}) yields

@@

We consider a two-layer atmosphere structure model as presented in section \ref{model}. We approximate the outer radiative zone of the atmosphere as nearly isothermal. Moreover, we do not take into account the self-gravity of the atmosphere in either region. We derive the structure of an isothermal non self-gravitating atmosphere is section \ref{iso}.

In the next subsections we derive the structure and cooling behavior of such an atmosphere, deferring some of the calculations to \App{sec:analytic}. We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}. Finally, we derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.

@@

We now consider the structure of a low mass (non self-gravitating) isothermal atmosphere.  We assume the atmosphere matches onto a constant background density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}). From equation (\ref{eq:structa}) and without taking into account the self-gravity of the atmosphere, the resulting density profile is
\begin{equation} \label{eq:rhoiso}
\rho = \rho_{\rm d} \exp \left({R_B \over r} - {1 \over n_{\rm{fit}}} \right) \approx   \rho_{\rm{d}} \exp \left(R_B \over r  \right),
\end{equation} 
where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.  However, the choice of boundary condition does have an order unity effect on the density near the Bondi radius. In our study, $n_{\rm{fit}}=R_{\rm H}/R_{\rm B}$.

The mass of the atmosphere is determined by integrating equation (\ref{eq:structb}) from the core to the Bondi radius using the density profile (\ref{eq:rhoiso}).  Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.  In this case
\begin{equation} \label{eq:MatmISO}
M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
\end{equation} 
This result is the leading order term in a series expansion.  Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.

@@

%\subsection{Temperature Contrast at Convective Boundary}
In this section we estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the radiative region is not purely isothermal. We show that the temperature contrast between the temperature at the radiative-convective boundary $T_{\rm RCB}$ and the ambient disk $T_{\rm d}$, is modest.  From equation (\ref{eq:delrad}), we express the radiative lapse rate
\begin{equation}\label{eq:delradan}
\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P_{\rm d})^{1+\alpha} \over (T/T_{\rm d})^{4-\beta}},
\end{equation}

\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}), and $\nabla_{\rm d}$ is the radiative temperature gradient at the disk:

\begin{equation}
\label{eq:delo}
\nabla \di \equiv \frac{3 \kappa(P{\di}, T{\di}) P{\di}}{64 \pi G M \sigma T_{\rm d}^4} L.
\end{equation}
